\section{Adding threshold queries to CvRDTs}\label{s:model}

In Shapiro \etal's CvRDT model,
the query operation $q$ reads the
exact contents of its local replica, and therefore different replicas
may see different states at the same time, if not all updates have
been propagated yet.  That is, it is possible to observe intermediate
states of a CvRDT replica.  Such intermediate observations are not
possible with
threshold queries. In this section, we show how to extend the CvRDT
model to accommodate threshold queries.

\subsection{Objects with threshold queries}

Definition~\ref{def:state-based-object-with-threshold-queries}
extends Shapiro \etal's definition of a state-based object with
a threshold query method $t$:
\begin{definition}[state-based object with threshold queries]
  \label{def:state-based-object-with-threshold-queries}
  A \emph{state-based object with threshold queries} (henceforth
  \emph{object}) is a tuple $(S, s^0, q, t, u, m)$, where $S$ is a set
  of states, $s^0 \in S$ is the initial state, $q$ is a \emph{query
    method}, $t$ is a \emph{threshold query
    method}, $u$ is an \emph{update method}, and $m$ is a \emph{merge
    method}.
\end{definition}

In order to give a semantics to the threshold query method $t$, we
need to formally define the notion of a threshold set described in
Section~\ref{s:threshold-reads}:

\begin{definition}[threshold set]
  \label{def:threshold-set}
  A \emph{threshold set with respect to a lattice $(S, \leq)$} is a
  set $\mathcal{S} = \{ S_a, S_b, \dots \}$ of one or more sets
  of \emph{activation states}, where each set of activation states is a subset of $S$,
  the set of
  lattice elements, and where the following \emph{pairwise
    incompatibility} property holds:

  For all $S_a, S_b \in \mathcal{S}$, if $S_a \neq S_b$, then for all
  activation states $s_a \in S_a$ and for all activation states $s_b \in
  S_b$, $s_a \sqcup s_b = \top$, where $\sqcup$ is the join operation
  induced by $\leq$ and $\top$ is the greatest element of $(S, \leq)$.
\end{definition}

In our model, we assume a finite set of $n$ processes $p_1, \dots, p_n$, and
consider a single replicated object with one replica at each process,
with replica $i$ at process $p_i$.  Processes may crash silently; we
say that a non-crashed process is \emph{correct}.

Every replica has initial state $s^0$.  Methods execute at individual
replicas, possibly updating that replica's state.  The $k$th method
execution at replica $i$ is written $f^{k}_{i}(a)$, where $k$ is $\geq 1$ and $f$ is
either $q$, $t$, $u$, or $m$, and $a$ is the arguments to $f$, if any.
Methods execute sequentially at each replica.
The state of replica $i$ after the $k$th method execution at $i$ is
$s^{k}_{i}$.  We say that states $s$ and $s'$ are equivalent, written
$s \equiv s'$, if $q(s) = q(s')$.

\subsection{Causal histories}

An object's \emph{causal history} is a record of all the updates that
 have happened at all replicas.  The causal history does
not track the order in which updates happened, merely that
they did happen.  The \emph{causal history at replica $i$ after
  execution $k$} is the set of all updates that have
happened at replica $i$ after execution $k$.
Definition~\ref{def:causal-history} updates Shapiro
\etal's definition of causal history for a state-based object
to account for $t$ (a trivial change, since execution of $t$
does not change a replica's causal history):
\begin{definition}[causal history]
  \label{def:causal-history}
  A \emph{causal history} is a sequence $[c_1, \dots, c_n]$, where
  $c_i$ is a set of the updates that have occurred at
  replica $i$.  Each $c_i$ is initially $\emptyset$.  If the $k$th
  method execution at replica $i$ is:
  \begin{itemize}
    \item a query $q$ or a threshold query $t$, then the causal
      history at replica $i$ after execution $k$ does not change:
      $c^{k}_{i} = c^{k-1}_{i}$.
    \item an update $u^{k}_{i}(a)$: then the causal history at replica
      $i$ after execution $k$ is $c^{k}_{i} = c^{k-1}_{i} \cup {
      u^{k}_{i}(a) }$.
    \item a merge $m^{k}_{i}(s^{k'}_{i'})$: then the causal history at
      replica $i$ after execution $k$ is the union of the local and
      remote histories: $c^{k}_{i} = c^{k-1}_{i} \cup c^{k'}_{i'}$.
  \end{itemize}
\end{definition}
We say that an update is \emph{delivered at replica $i$} if it is in
the causal history at replica $i$.

\subsection{Threshold CvRDTs and the semantics of blocking}\label{ss:blocking}

With the previous definitions in place, we
can give the definition of a CvRDT that supports threshold queries:

\lk{OK, it's a little confusing to use the letter $t$ for the
  threshold query method, because op-based CRDTs use $t$ for the
  ``prepare-update'' method. But, since this paper doesn't have
  op-based CRDTs, there's no ambiguity...I'm stealing it.}

\begin{definition}[CvRDT with threshold queries]
  \label{def:cvrdt-with-threshold-queries}
  A \emph{convergent replicated data type with threshold queries}
  (henceforth \emph{threshold CvRDT}) is an object equipped
  with a partial order $\leq$, written $(S, \leq, s^0, q, t, u, m)$,
  that has the following properties:
  \begin{itemize}
    \item $S$ forms a join-semilattice ordered by $\leq$.
    \item $S$ has a greatest element $\top$ according to $\leq$.
    \item The query method $q$ takes no arguments and returns the
      local state.
    \item The threshold query method $t$ takes a threshold set
      $\mathcal{S}$ as its argument, and has the following semantics:
      let $t^{k+1}_i(\mathcal{S})$ be the $k+1$th method execution at
      replica $i$, where $k \geq 0$.  If, for some activation state
      $s_a$ in some (unique) set of activation states $S_a \in
      \mathcal{S}$, the condition $s_a \leq s^{k}_{i}$ is met,
      $t^{k+1}_i(\mathcal{S})$ returns the set of activation states
      $S_a$.  Otherwise, $t^{k+1}_i(\mathcal{S})$ returns the
      distinguished value $\block$.
    \item The update method $u$ takes a state as argument and updates
      the local state to it.
\rn{Wait updates the local state to it?  Or to the LUB of the input
  state and it?  And if it's the latter, then do we need the next
  bullet, or is the inflationary property a consequence of this bullet?}
\lk{It actually doesn't do a lub, it just does an inflationary update
  of any kind.  the only lub that happens is with the merge method.}
    \item State is \emph{inflationary} across updates: if $u$ updates
      a state $s$ to $s'$, then $s \leq s'$.
    \item The merge method $m$ takes a remote state as its argument,
      computes the join of the remote state and the local state with
      respect to $\leq$, and updates the local state to the result.
  \end{itemize}
  and the $q$, $t$, $u$, and $m$ methods have no side effects other
  than those listed above.
\end{definition}
We use the $\block$ return value to model $t$'s ``blocking'' behavior
as a mathematical function with no intrinsic notion of running
duration.  When we say that a call to $t$ ``blocks'', we mean that it
immediately returns $\block$, and when we say that a call to $t$
``unblocks'', we mean that it returns a set of activation states
$S_a$.

Modeling blocking as a distinguished value introduces a new
complication: we lose determinism, because a call to $t$ at a
particular replica may return either $\block$ or a set of activation
states $S_a$, depending on the replica's state at the time it is
called.  However, we can conceal this nondeterminism with an
additional layer over the nondeterministic API exposed by $t$.  This
additional layer simply \emph{polls} $t$, calling it repeatedly until
it returns a value other than $\block$.  Calls to $t$ at a replica
that are made by this ``polling layer'' count as method executions at
that replica, and are arbitrarily interleaved with other method
executions at the replica, including updates and merges.  The polling
layer itself need not do any computation other than checking to see
whether $t$ returns $\block$ or something else; in particular, the
polling layer does not need to compare activation states to replica
states, since that comparison is done by $t$ itself.  
% Indeed, a
% polling mechanism like this is how threshold reads for LVars are
% implemented in practice~\cite{Freeze-paper}.
\rn{I'm afraid the prev sentence about polling/lvish gives the wrong
  impression, and a more nuanced one would take more space.  So how
  about we nix it?  The problem is that LVish is of course event
  driven.  It doesn't sit around polling gets waiting for puts to
  happen.  It only does anything when the put happens.  And yet, the
  mapping of puts onto exact callbacks is not always perfectly
  precise.  With an IVar it would be, but with an ``IMap'', the
  there's a callback registered on a particular key, and yet it might
  be polled every time a write is made to ANY key.  That's just a
  limitation of the implementation currently -- it can't do finer
  grain lists of guarded callbacks.  More generally, some lvars
  really do *need* polling, yet they only need to poll when a write
  occurs.}

The set of activation states $S_a$ that a call to $t$ returns when it
unblocks is unique because of the pairwise incompatibility property
described in Definition~\ref{def:threshold-set}.  The intuition behind
pairwise incompatibility is that described in
Section~\ref{subsec:pairwise-incompatibility}: without it, different
orderings of updates could allow the same threshold query to unblock
in different ways, introducing nondeterminism that would be observable
beyond the polling layer.

\subsection{Threshold CvRDTs are strongly eventually consistent}

\lk{Hmmm.  We don't have to do all this blathering about termination
  anymore, because now threshold queries actually do always return
  immediately!}

We can define eventual consistency and strong eventual consistency
exactly as Shapiro \etal~do in their model.  In the following
definitions, a \emph{correct replica} is a replica at a correct
process, and the  symbol $\eventually$ means ``eventually'':

%% Since Shapiro~\etal's CvRDTs are strongly eventually
%% consistent~\cite{crdts}, do we have the same guarantee for threshold
%% CvRDTs?
%% The addition of threshold queries to the model introduces a wrinkle
%% with regard to termination.  Recall from Section~\ref{s:cvrdts} that
%% Shapiro~\etal's definition of eventual consistency (and therefore also
%% their definition of strong eventual consistency) includes a
%% \emph{termination} property that says that all method executions
%% terminate.  But a threshold query could block forever, if no update
%% arrives to unblock it.  The determinism of threshold queries does \emph{not} rule out
%% queries that block forever.  Rather, it says that if a threshold query blocks
%% forever on one execution, then it will do so on \emph{every}
%% execution, regardless of scheduling.
%% For that to be the case, we need to ensure that
%% blocking threshold queries do not race with merges that could unblock
%% them.  We accomplish this by modeling each process as two
%% \emph{threads}, the \emph{main thread} and the \emph{merge thread}.
%% Updates ($u$), ordinary queries ($q$), and threshold queries ($t$) take place on the
%% main thread; merges ($m$) take place on the merge thread and are interleaved
%% with operations on the main thread.  Hence a merge operation cannot be
%% blocked behind a threshold query.

%% We update Shapiro \etal's definitions of eventual
%% consistency and strong eventual consistency to account for threshold
%% queries not necessarily terminating:

\begin{definition}[eventual consistency (EC)]
  \label{def:eventual-consistency}
  An object is \emph{eventually consistent} (EC) if the following three
  conditions hold:
  \begin{itemize}
    \item \emph{Eventual delivery}: An update delivered at some
      correct replica is eventually delivered at all correct replicas:
      $\forall i, j : f \in c_i \implies \eventually f \in c_j$.
    \item \emph{Convergence}: Correct replicas at which the same
      updates have been delivered eventually have equivalent state:
      $\forall i, j : c_i = c_j \implies \eventually s_i \equiv s_j$.
    \item \emph{Termination}: All method executions halt.
  \end{itemize}
\end{definition}

\begin{definition}[strong eventual consistency (SEC)]
  \label{def:strong-eventual-consistency}
  An object is \emph{strongly eventually consistent} (SEC) if it is
  eventually consistent and the following condition holds:
  \begin{itemize}
  \item \emph{Strong convergence}: Correct replicas at which the same
    updates have been delivered have equivalent state:
    $\forall i, j : c_i = c_j \implies s_i \equiv s_j$.
  \end{itemize}
\end{definition}
Since we model blocking threshold queries with $\block$, we need not
be concerned with threshold queries not necessarily terminating.
Determinism does \emph{not} rule out queries that return $\block$
every time they are called (and would therefore cause the polling
layer to block forever).  However, we guarantee that if a threshold
query returns $\block$ every time it is called during a complete run
of the system, it will do so on \emph{every} run of the system,
regardless of scheduling.  That is, it is not possible for a query to
cause the polling layer to block forever on some runs, but not on
others.\lk{I changed the wording here from ``execution'' to ``complete
  run of the system'' to lessen confusion with individual method
  executions.}

Finally, we can directly leverage Shapiro 
\etal's SEC result for CvRDTs to show that a threshold CvRDT is SEC:
\begin{theorem}[strong eventual consistency of threshold CvRDTs]
  \label{thm:strong-eventual-consistency-of-threshold-cvrdts}
  Assuming eventual delivery and termination, an object that meets
  the criteria for a threshold CvRDT is SEC.
\end{theorem}
\begin{proof}
From Shapiro \etal, we have that an object that meets the criteria for
a CvRDT is SEC~\cite{crdts}.  Shapiro \etal's proof also assumes that
eventual delivery and termination hold for the object, and proves that
strong convergence holds --- that is, that given causal histories
$c_i$ and $c_j$ for respective replicas $i$ and $j$, that their states
$s_i$ and $s_j$ are equivalent.  The proof relies on the commutativity
of the least upper bound operation.  Since, according to our
Definition~\ref{def:causal-history}, threshold queries do not affect
causal history, we can leverage Shapiro \etal's result to say that a
threshold CvRDT is also SEC.
\qed
\end{proof}

\rn{Do we not need to grapple with the blocking forever behavior here
  with respect to reusing this result off the shelf?}
\lk{right, we don't, because we aren't proving anything about the
  polling layer, just the inner CvRDT layer.}

