\begin{abstract}

  Programs written using a {\em deterministic-by-construction} model
  of parallel computation are guaranteed to always produce the same
  observable results, offering programmers freedom from subtle,
  hard-to-reproduce nondeterministic bugs that are the scourge of
  parallel software.
  We present \emph{LVars}, a new model for deterministic-by-construction parallel
  programming that generalizes existing single-assignment models to
  allow multiple assignments that are monotonically increasing with
  respect to a user-specified lattice.  
  LVars ensure determinism by
  allowing only monotonic writes and ``threshold'' reads that block
  until a lower bound is reached.  We give a proof of determinism
  and a prototype implementation
  for a language with LVars
  and describe how to extend the LVars model to support a limited form of
  nondeterminism that admits failures but never wrong answers.
%   \rn{I'm a bit worried about saying ``both those with shared state''
%     since shared write-once state is quite a bit different than what
%     most people usually think of.}

\if{0}
  There is a need for new proposals that add diversity to the current
  menu of determistic parallel programming options.
  This paper presents a variant of the $\lambda$-calculus, $\lambdapar$: a
  parallel, call-by-value calculus that includes shared mutable
  variables whose states occupy a join semilattice and move
  monotonically upwards in that lattice.  
  In this paper we prove $\lambdapar$ deterministic.  
  Further, we explain how $\lambdapar$ generalizes a number of
  existing deterministic parallel programming models, and provides a
  new foundation for exploring {limited} forms of nondeterminism.
\fi{}

\end{abstract}
