Not long ago, someone asked me if knew I wanted to work on LVars at
the time I applied to grad school, and I had to laugh.  Not only did I
not know then that I wanted to work on LVars, I didn't even know what
subfield of computer science I wanted to study.  I had an idea of what
kinds of problems I thought were interesting, but I didn't really
grasp that there was actually a field of study devoted to those
problems and that ``programming languages'' was what it was called.
As it happened, one of the Ph.D. programs to which I applied turned
out to be a very good place to study PL; it also so happened that that
program was the only one that accepted me.  So the first
acknowledgement I want to make here is to the Indiana University
computer science program for taking a chance on me and making this
dissertation possible.

I took Dan Friedman's programming languages course in my first
semester at IU in fall 2008, and the following spring, Dan invited me
to be a teaching assistant for the undergraduate edition of the
course, known as C311, together with his student and collaborator Will
Byrd.  This led to my first PL research experience, which consisted of
playing with a fantastic little language called
miniKanren\footnote{\url{http://www.minikanren.org} --- not that any
such website existed at the time!} during the summer of 2009 with Dan,
Will, and a group of relational programming aficionados fueled by
intense curiosity and Mother Bear's pizza.  At the time, I was utterly
unprepared to do research and was in way over my head, but we had a
lot of fun, and I went on to help out with C311 and another course,
H211, for three more semesters under Dan and Will's expert guidance.

Amal Ahmed joined the IU faculty in 2009 and was a tremendous
influence on me as I began truly learning how to read research papers
and, eventually, write them.  Although Amal and I never worked on
determinism as such, it was nevertheless from Amal that I learned how
to design and prove properties of calculi much like the $\lambdaLVar$
and $\lambdaLVish$ calculi in this dissertation.  Amal also had a huge
influence on the culture of the PL group at IU.  Weekly ``PL Wonks''
meetings are a given these days, but it wasn't so very long ago that
the PL group didn't have regular meetings or even much of a group
identity.  It was Amal who rounded us all up and told us that from
then on, we were all to give talks every semester.  (Happily, this
process has now become entirely student-run and self-sustaining.)  It
was also at Amal's urging that I attended my first PL conference, ICFP
2010 in Baltimore, which was where I first met a surprising number of
the people who have since become my mentors, colleagues,
collaborators, and dear friends in the PL community, including Neel
Krishnaswami, Chung-chieh Shan, Tim Chevalier, Rob Simmons, Jason
Reed, Ron Garcia, Stevie Strickland, Dan Licata, and Chris Martens.

When Amal left IU for Northeastern in 2011, I had to make the
difficult decision between going with her or staying, and although I
chose to stay at IU and work with Ryan Newton on what would eventually
become this dissertation, I think that Amal's influence on the flavor
of the work I ended up doing is evident.  (In fact, it was Amal who
originally pointed out the similarity between the Independence Lemma
and the frame rule that I discuss in
Section~\ref{subsection:lvars-independence}.)

Ryan Newton, my advisor, runs at a faster clock speed than most
people.  His first act as a new faculty member at Indiana in fall 2011
was to arrange for our building to get a fancy espresso machine ---
and to kick-start a community of student espresso enthusiasts who
lovingly maintain it.  In fact, now that I think about it, I'm not
sure it was a coincidence that Ryan had the espresso machine installed
in the kitchen directly next to my office\dots Anyway, I took Ryan's
seminar course that fall, which was how I started learning about
deterministic parallel programming models.  In November 2011, we began
enthusiastically discussing the idea of generalizing single-assignment
models, and Ryan suggested that I could help him write a grant
proposal to continue working on the idea.  We submitted our proposal
in December 2011, and to my astonishment, it was funded on our first
try.  I gratefully acknowledge everyone at the National Science
Foundation who was involved in the decision to fund grant CCF-1218375,
since without that early vote of confidence, I'm not sure I would have
been able to keep my spirits up during the year that followed, in
which it took us four attempts to get our first LVars paper published.
(And, although I didn't feel thankful at the time, I'm now also
thankful to the anonymous reviewers of POPL 2013, ESOP 2013, and ICFP
2013, whose constructive criticism helped us turn LVars from a
half-baked idea into a convincing research contribution.)

In addition to Ryan, I was lucky to have a wonderful dissertation
committee consisting of Amr Sabry, Larry Moss, and Chung-chieh Shan.
Amr, Larry, and Ken all played an important part early in the
development of LVars and helped resolve subtle issues with the
semantics of $\lambdaLVar$ and its determinism proof.  What appears
now in Chapter~\ref{ch:lvars} is relatively simple --- but it is only
simple because of considerable effort spent making it so!  I'm also
grateful to all my committee members for showing up to my thesis
proposal at eight-thirty in the morning in the middle of an Indiana
December snowstorm.  Additionally, I want to offer special thanks to
Sam Tobin-Hochstadt, Neel Krishnaswami, and Aaron Turon --- all of
whom gave me so much guidance, encouragement, and support throughout
this project that I think of them as \emph{de facto} committee
members, too.

By fall 2012, working on LVars had gotten me interested in the idea of
developing a separation logic for deterministic parallelism.  I
reached out to Aaron and Neel, and they were enthusiastic about
collaborating with Ryan and me.  When I met up with them in
Saarbr{\"u}cken in January 2013, though, they suggested that we first
focus on making the LVars programming model more expressive, and we
put the separation logic idea on hold while we went on an exciting
side trip into language design.  That ``side trip'' eventually took us
all the way to event handlers, quiescence, freezing, and
quasi-determinism, the topics of Chapter~\ref{ch:quasi}, and to our
POPL 2014 paper.  It also led to the LVish Haskell library, the topic
of Chapter~\ref{ch:lvish}, which was largely implemented by Aaron and
Ryan.  I did the proof of quasi-determinism for $\lambdaLVish$ in
collaboration with Neel.  I'm thankful also to Derek Dreyer for
helping facilitate our collaboration, as well as for giving me an
excuse to give a \emph{Big Lebowski}-themed talk at MPI-SWS.

One of the most exciting and rewarding parts of my dissertation work
has been pursuing the connection between LVars and distributed data
consistency --- the topic of Chapter~\ref{ch:distributed}.  It's often
a challenge for people coming from different subfields to find common
ground, and I'm indebted to the many distributed systems researchers
and practitioners who have met me more than halfway, particularly the
BOOM group at Berkeley and the Riak folks at Basho.  Speaking at RICON
2013 about LVars was one of the highlights of my Ph.D. experience;
thanks to everyone who made it so.

Katerina Barone-Adesi, Jos\'{e} Valim, and Zach Allaun --- not
coincidentally, all members of the Recurse Center community, which is
the best programming community in the world --- gave feedback on
drafts of this dissertation and helped me improve the presentation.
Thanks to all of them.

Jason Reed contributed the wonderful spot illustrations that appear
throughout.  My hope is that they'll make the pages a bit more
inviting and offer some comic relief to the frustrated reader.
They've certainly done that for me.

Finally, this dissertation is dedicated to my amazing partner, Alex
Rudnick, without whose love, support, advice, and encouragement I
would never have \emph{started} grad school, let alone finished.
