% A tweaked version of my CV, just for dissertation purposes.  The
% only differences between this and my standard CV are:
% - no page headers
% - no page numbering
% - black font everywhere
% - no date footer on the last page

\documentclass[10pt,letterpaper]{article}

\usepackage{hyperref}
\usepackage{geometry}
\usepackage{amssymb}
\usepackage{xcolor}

% Font stuff.
\usepackage{xltxtra,fontspec,xunicode}
\defaultfontfeatures{Scale=MatchLowercase}
\setromanfont[Mapping=tex-text]{PT Serif}
\setsansfont[Mapping=tex-text]{PT Serif}

% Name.
\def\name{Lindsey Kuper}

% The following metadata will show up in the PDF properties.
\hypersetup{
  colorlinks = true,
  urlcolor = black,
  pdfauthor = {\name},
  pdftitle = {\name: Curriculum Vitae},
  pdfsubject = {Curriculum Vitae},
  pdfpagemode = UseNone
}
\urlstyle{same}

\geometry{
  body={6.6in, 9.0in},
  left=0.95in,
  top=1.0in
}

% Custom section fonts.
\usepackage{sectsty}
\sectionfont{\mdseries\Large}
\subsectionfont{\mdseries\itshape\large}

% Don't indent paragraphs.
\setlength\parindent{0em}

% Suppress page numbers.
\pagestyle{empty}

% One way to get a small square bullet.
\newcommand{\localtextbulletone}{\textcolor{gray}{\raisebox{.45ex}{\rule{.6ex}{.6ex}}}}
\renewcommand{\labelitemii}{\localtextbulletone}
\renewcommand{\labelitemi}{\localtextbulletone}

\newenvironment{itemize*}
  {\begin{itemize}
      % An opportunity to tweak spacing.
      \setlength{\itemsep}{1pt}
      \setlength{\parskip}{3pt}
  }
  {\end{itemize}}

\begin{document}

\centerline{\Huge \textbf{\name}}

\bigskip
\bigskip

\begin{minipage}[t]{0.5\textwidth}
  \small
  Programming Systems Lab, \href{http://www.intel.com/content/www/us/en/research/intel-research.html}{Intel Labs} \\
  Intel Corporation \\
  3600 Juliette Lane \\
  Santa Clara, CA 95054
\end{minipage}
\begin{minipage}[t]{0.5\textwidth}
  \small
  \raggedleft
  \href{mailto:lindsey.kuper@intel.com}{\nolinkurl{lindsey.kuper@intel.com}} \\
  \href{mailto:lindsey@composition.al}{\nolinkurl{lindsey@composition.al}} \\ 
  \url{http://www.cs.indiana.edu/~lkuper} \\
  \url{http://composition.al}
\end{minipage}

\section*{Education}

\begin{itemize*}
\item
\textbf{Indiana University}, School of Informatics and Computing, 2008--2015
\begin{itemize*}
\item Ph.D. Computer Science, September 2015. \\
      Research committee: Ryan R. Newton (chair), Lawrence S. Moss,
      Amr Sabry, Chung-chieh Shan. \\
      Dissertation: \emph{Lattice-based Data Structures for
        Deterministic Parallel and Distributed Programming}.

\item M.S. Computer Science, May 2010.
\end{itemize*}

\item
\textbf{Grinnell College}, 2000--2004
\begin{itemize*}
\item B.A. Computer Science and Music (with honors), May 2004.
\end{itemize*}

\item Additional coursework and summer schools:
\begin{itemize*}
\item
  \href{http://www.cs.uoregon.edu/research/summerschool/summer13/}{Oregon
    Programming Languages Summer School: Types, Logic, and
    Verification}, Summer 2013.
\item
  \href{http://www.cs.uoregon.edu/research/summerschool/summer12/}{Oregon
    Programming Languages Summer School: Logic, Languages,
    Compilation, and Verification}, Summer 2012.
\item Operating Systems, Cornell University, Summer 2010.
\end{itemize*}

\end{itemize*}

\section*{Employment history}

\begin{itemize*}
\item{\textbf{Intel Corporation (Programming Systems Lab, \href{http://www.intel.com/content/www/us/en/research/intel-research.html}{Intel Labs})}, Santa Clara, CA}
  \begin{itemize*}
  \item \textit{Research Scientist}, September 2014--present.
  \end{itemize*}

\item{\textbf{Indiana University}, Bloomington, IN}
  \begin{itemize*}
  \item \textit{Research Assistant} (with Ryan Newton), January 2012--August 2014.
  \item \textit{Associate Instructor} (with William E. Byrd), August 2011--December 2011.
  \item \textit{Research Assistant} (with Amal Ahmed), August 2010--December 2010.
  \item \textit{Associate Instructor} (with Daniel P. Friedman), January 2009--May 2010.
  \end{itemize*}

\item \textbf{\href{http://www.mozilla.org}{Mozilla Corporation}},
  Mountain View, CA
  \begin{itemize*}
  \item \textit{Research Engineering Intern} (supervised by Brian Anderson, Niko Matsakis and Patrick Walton), May--August 2012.
  \item \textit{Research Engineering Intern} (supervised by Dave Herman), March--August 2011.
  \end{itemize*}

\item \textbf{\href{http://www.grammatech.com}{GrammaTech, Inc.}},
  Ithaca, NY
  \begin{itemize*}
  \item \textit{Software Engineering Intern} (supervised by David Melski), May--August 2010.
  \end{itemize*}

\item \textbf{\href{http://www.bfwpub.com}{Bedford, Freeman and Worth
      Publishing Group}}, New York, NY and Portland, OR
  \begin{itemize*}
  \item \textit{Associate Project Manager}, July 2006--June 2008.
  \end{itemize*}

\item \textbf{IBCTV, LLC}, Chicago, IL and Portland, OR
  \begin{itemize*}
  \item \textit{Web Designer/Developer}, August 2004--June 2006.
  \end{itemize*}

%% \item \textbf{\href{http://www.internaldrive.com/}{internalDrive,
%%       Inc.}}, Austin, TX and Evanston, IL
%%   \begin{itemize*}
%%   \item \textit{Instructor}, June--August 2004.
%%   \item \textit{Instructor}, June--August 2003.
%%   \end{itemize*}
\end{itemize*}

\section*{Research funding}

\begin{itemize*}
\item Co-wrote (with Ryan Newton) NSF grant CCF-1218375,
  \href{http://nsf.gov/awardsearch/showAward?AWD_ID=1218375}{Generalizing
    Monotonic Data Structures for Expressive, \\
    Deterministic Parallel Programming} (\$377,315; 8/1/2012--7/31/2015),
  which funded my dissertation work.
\end{itemize*}

\section*{Conference publications}

\begin{itemize*}
\item
  \textbf{Lindsey Kuper}, Aaron Todd, Sam Tobin-Hochstadt and Ryan
  R. Newton. \\
  \href{http://www.cs.indiana.edu/~lkuper/papers/effectzoo-pldi14.pdf}{Taming
    the parallel effect zoo: extensible deterministic parallelism with
    LVish.} \\
  In \emph{Proceedings of the
    \href{http://conferences.inf.ed.ac.uk/pldi2014/}{35th ACM SIGPLAN
      Conference on Programming Language Design and Implementation
      (PLDI '14)}}, \\
  Edinburgh, UK, June 2014. \\
  (52/287 $\approx$ 18\% accepted)

\item
  \textbf{Lindsey Kuper}, Aaron Turon, Neelakantan R. Krishnaswami and
  Ryan R. Newton. \\
  \href{http://www.cs.indiana.edu/~lkuper/papers/lvish-popl14.pdf}{Freeze
    after writing: quasi-deterministic parallel programming with
    LVars.} \\
  In \emph{Proceedings of the
    \href{http://popl.mpi-sws.org/2014/}{41st ACM SIGPLAN-SIGACT
      Symposium on Principles of Programming Languages (POPL '14)}}, \\
  San Diego, CA, January 2014. \\
  (51/220 $\approx$ 23\% accepted)
\end{itemize*}

\section*{Workshop publications}
\begin{itemize*}
\item
  \textbf{Lindsey Kuper} and Ryan R. Newton. \\
  \href{http://www.cs.indiana.edu/~lkuper/papers/joining-wodet14.pdf}{Joining
    forces: toward a unified account of LVars and convergent
    replicated data types.} \\
  In the \emph{\href{http://wodet.cs.washington.edu/}{5th Workshop on
      Determinism and Correctness in Parallel Programming (WoDet
      '14)}}, \\
  Salt Lake City, UT, March 2014.
  % (9/11 $\approx$ 82\% accepted)

\item
  \textbf{Lindsey Kuper} and Ryan R. Newton. \\
  \href{http://www.cs.indiana.edu/~lkuper/papers/lvars-fhpc13.pdf}{LVars:
    lattice-based data structures for deterministic parallelism.} \\
  In \emph{Proceedings of the
    \href{http://hiperfit.dk/fhpc13.html}{2nd ACM SIGPLAN Workshop on
      Functional High-Performance Computing (FHPC '13)}}, \\
  Boston, MA, September 2013.
  %% (8/14 $\approx$ 57\% accepted)

\item
  Andrew W. Keep, Michael D. Adams, \textbf{Lindsey Kuper}, William
  E. Byrd and Daniel P. Friedman. \\
  \href{http://www.cs.indiana.edu/~lkuper/papers/lambdae.pdf}{A
    pattern matcher for miniKanren, or, how to get into trouble with
    CPS macros.}  \\
  In \emph{Proceedings of the
    \href{http://www.schemeworkshop.org/2009/}{2009 Scheme and
      Functional Programming Workshop (Scheme '09)}}, \\
  Boston, MA, August 2009.
\end{itemize*}

\section*{Technical reports}

\begin{itemize*}
\item
  \textbf{Lindsey Kuper}, Aaron Turon, Neelakantan R. Krishnaswami
  and Ryan R. Newton. \\
  \href{http://www.cs.indiana.edu/cgi-bin/techreports/TRNNN.cgi?trnum=TR710}{Freeze
    after writing: quasi-deterministic parallel programming with LVars.} (56 pages) \\
  Indiana University Technical Report TR710, November 2013.

\item 
  \textbf{Lindsey Kuper} and Ryan R. Newton. \\
  \href{http://www.cs.indiana.edu/cgi-bin/techreports/TRNNN.cgi?trnum=TR702}{A
    lattice-theoretical approach to deterministic parallelism with
    shared state.} (60 pages) \\
  Indiana University Technical Report TR702, October 2012.

\item David Melski, David Cok, John Phillips, Scott Wisniewski, Suan
  Hsi Yong, Nathan Lloyd, \textbf{Lindsey Kuper}, Denis Gopan and
  Alexey Loginov. \\
  \href{http://www.dtic.mil/dtic/tr/fulltext/u2/a532995.pdf}{Safety in
    numbers.} (104 pages) \\
  % http://www.dtic.mil/docs/citations/ADA532995
  GrammaTech, Inc. project final report, November 2010.
\end{itemize*}

\newpage

\section*{Talks}

\begin{itemize*}
\item
  \emph{LVars for distributed programming, or, LVars and CRDTs join forces.}
  \begin{itemize*}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/wg28-may-2015.pdf}{IFIP Working Group 2.8 (Functional Programming), Kefalonia, Greece, May 26, 2015.}
  \end{itemize*}
\item
  \emph{LVars: lattice-based data structures for deterministic
    parallel and distributed programming.}
  \begin{itemize*}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/compose-2015.pdf}{Compose :: Conference, New York, NY, January 31, 2015.}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/hacker-school-march-2014.pdf}{Hacker
      School, New York, NY, March 24, 2014.}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/intel-labs-march-2014.pdf}{Intel
      Labs, Santa Clara, CA, March 21, 2014.}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/utah-march-2014.pdf}{University
      of Utah, Salt Lake City, UT, March 4, 2014.}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/msr-january-2014.pdf}{Microsoft
      Research, Mountain View, CA, January 27, 2014.}
  \end{itemize*}

\item
  \emph{Joining forces: toward a unified account of LVars and convergent replicated data types.}
  \begin{itemize*}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/wodet-march-2014.pdf}{WoDet
      2014, Salt Lake City, UT, March 2, 2014.}
  \end{itemize*}

\item
  \emph{Freeze after writing: quasi-deterministic parallel programming
    with LVars.}
  \begin{itemize*}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/lvish-popl-2014.pdf}{POPL
      2014, San Diego, CA, January 23, 2014.}
  \end{itemize*}

\item
  \emph{LVars: lattice-based data structures for deterministic parallelism.}
  \begin{itemize*}
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/lvars-mozilla-2013.pdf}{Mozilla
        Corporation, Mountain View, CA, October 31, 2013.}
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/ricon-2013.pdf}{RICON
        West 2013, San Francisco, CA, October 29, 2013.}
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/fhpc-2013.pdf}{FHPC
        2013, Boston, MA, September 23, 2013.}
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/hacker-school-june-2013.pdf}{Hacker
        School, New York, NY, June 10, 2013.}
    \end{itemize*}

%% \item
%%   \emph{LVars: deterministic parallelism through shared monotonic data
%%     structures.}
%%   \begin{itemize*}
%%     \item
%%       \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/cnc-april-2013.pdf}{Open
%%         CnC Meeting, April 17, 2013.}
%%     \end{itemize*}

\item
  \emph{A lattice-based approach to deterministic parallelism.}
  \begin{itemize*}
    \item \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/mpi-sws-january-2013.pdf}{MPI-SWS,
      Saarbr\"{u}cken, Germany, January 30, 2013.}

    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/popl-student-session-2013.pdf}{POPL
        2013 student talk session, Rome, Italy, January 25, 2013.}
  \end{itemize*}

%% \item
%%   \emph{You gotta keep 'em separated: frame rules for reasoning about
%%     stateful and/or concurrent programs.}
%%   \begin{itemize*}
%%     \item
%%       \href{http://www.math.indiana.edu/seminars/event.phtml?id=3547}{IU Logic Seminar, November 28, 2012}
%%   \end{itemize*}

\item
  \emph{A lattice-based approach to deterministic parallelism with
    shared state.}
  \begin{itemize*}
    %% \item IU PL Colloquium Series, October 5, 2012.
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/monotonic-determinism-aarhus-2012.pdf}{Aarhus
        University, Aarhus, Denmark, September 14, 2012.}
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/monotonic-determinism/monotonic-determinism-berkeley-2012.pdf}{University
        of California--Berkeley, Berkeley, CA, August 16, 2012.}
  \end{itemize*}

\item \emph{Rust typeclasses turn trait-er.}
  \begin{itemize*}
  \item \href{http://www.cs.indiana.edu/~lkuper/talks/rust-traits/typeclasses-turn-trait-er-mozilla.pdf}{Mozilla
    Corporation, Mountain View, CA, August 9, 2012.}
  \end{itemize*}

\item \emph{Hacking the Rust object system at Mozilla.}
  \begin{itemize*}
  \item
    \href{http://www.cs.indiana.edu/~lkuper/talks/rust-objects/rust-objects-grinnell.pdf}{Grinnell
      College, Grinnell, IA, April 5, 2012.} \\ 
    (invited talk, hosted by the Grinnell
    \href{http://www.grinnell.edu/offices/dean/alumscholars/}{Alumni
      Scholars Program})
  \end{itemize*}

%% \item \emph{Monotonic data structures as a guiding principle for deterministic parallel programming}.
%%   \begin{itemize*}
%%   \item IU PL Colloquium Series, March 30, 2012.
%%   \end{itemize*}
%% \item \emph{A loopy introduction to dependence analysis}.
%%   \begin{itemize*}
%%   \item IU PL Colloquium Series, December 9, 2011.
%%   \end{itemize*}

\item
  \emph{Some pieces of the Rust object system: extension, overriding,
    and self.}
  \begin{itemize*}
    %% \item
    %%   \href{http://www.cs.indiana.edu/~lkuper/talks/rust-objects/rust-objects-iu.pdf}{
    %%     IU PL Colloquium Series, September 9, 2011.}
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/rust-objects/rust-objects.pdf}{Mozilla
        Corporation, Mountain View, CA, August 18, 2011.}
  \end{itemize*}

\item
  \emph{Parametric
      polymorphism through run-time sealing, or, theorems for low, low
      prices!}
  \begin{itemize*}
    \item
      \href{http://www.cs.indiana.edu/~lkuper/talks/multilang-param/multilang-param-northeastern.pdf}{Northeastern University, Boston, MA, February 23,
        2011.}
    %% \item IU PL Colloquium Series, February 18, 2011.
  \end{itemize*}

\item \emph{A system for testing specifications of CPU semantics, or,
  what I did on my summer vacation}.
  \begin{itemize*}
    %% \item IU PL Colloquium Series, October 8, 2010.
    \item \href{http://www.cs.indiana.edu/~lkuper/talks/tsl-validator/tsl-validator.pdf}{GrammaTech, Inc., Ithaca, NY, August 20, 2010.}
  \end{itemize*}

%% \item \emph{Efficient representations of triangular substitutions: a
%%   comparison in miniKanren}.
%%   \begin{itemize*}
%%   \item IU PL Colloquium Series, January 15, 2010.
%%   \end{itemize*}

\end{itemize*}

\section*{Undergraduate projects advised}
\begin{itemize*}
\item Isaiah Weating, Indiana University, Spring 2013.  Project title:
  \emph{Parallel Programming with LVars}.  Awarded third place in IU
  Undergraduate Research Opportunities in Computing (UROC) Poster
  Competition, May 2013.
\end{itemize*}

\section*{Teaching}
\begin{itemize*}
\item{\textbf{Associate Instructor, Indiana University}}
\begin{itemize*}
\item Fall 2011: \href{http://www.cs.indiana.edu/cgi-pub/h211/}{CSCI
  H211 Introduction to Computer Science, Honors}, taught by Will Byrd.
  Taught labs on Scheme programming and Arduino development, with a
  focus on procedural music generation.  Graded homework assignments
  and exams and held office hours.

\item Spring 2009, Fall 2009, Spring 2010:
  \href{http://www.cs.indiana.edu/classes/b521}{CSCI B521 Programming
    Language Principles} and
  \href{http://www.cs.indiana.edu/classes/c311}{CSCI C311 Programming
    Languages}, taught by Dan Friedman.  Taught lab sections, graded
  homework assignments and exams, and held office hours.  Course
  topics included environment-passing and continuation-passing
  interpreters; continuation-passing style, trampolining,
  registerization, and other correctness-preserving transformations;
  hygienic macro expansion; types and type inference; and functional
  and logic programming.  Nominated by students for 2009--2010
  Associate Instructor of the Year award.

\end{itemize*}
\item{\textbf{Instructor, internalDrive, Inc.}}

  Taught short project-based courses to middle-school and high-school
  students in a university setting.
\begin{itemize*}
\item Summer 2004 (Northwestern University): Various week-long courses
  on digital music editing and web design.

\item Summer 2003 (UT Austin): Various week-long courses on digital
  music editing, web design, and stop-motion animation.
\end{itemize*}
\end{itemize*}

\section*{Service}

\subsection*{Research community service}
\begin{itemize*}
\item Program chair:
  \begin{itemize*}
  \item
    \href{http://conf.researchr.org/track/POPL-2016/OBT-2016-talks}{Off
      the Beaten Track (OBT) 2016}.
  \end{itemize*}

\item Program committees:
  \begin{itemize*}
  \item
    \href{http://2016.ecoop.org/committee/ecoop-2016-papers-external-review-committee}{ECOOP
      2016}, external review committee.

  \item
    \href{http://conf.researchr.org/committee/POPL-2016/}{POPL 2016}, external review
    committee.

  \item \href{http://ifl2015.wikidot.com/}{IFL 2015}.

  \item
    \href{http://2015.splashcon.org/track/onward2015-papers}{Onward!
      Papers 2015}.

  \item \href{http://papoc.di.uminho.pt}{Workshop on Principles and
    Practice of Consistency for Distributed Data (PaPoC) 2015}.

  \item \href{http://www.cs.rice.edu/~sc40/obt15/}{Off the Beaten
    Track (OBT) 2015}.

  \item \href{http://ifl2014.github.io/}{IFL 2014}.

  \item \href{http://www.haskell.org/haskell-symposium/2014/}{Haskell
    Symposium 2014}.
  \end{itemize*}

\item Journal reviewing:
  \href{http://link.springer.com/journal/446}{Distributed Computing}, 2015;
  \href{http://compilers.cs.ucla.edu/toplas/}{ACM Transactions on
    Programming Languages and Systems (TOPLAS)}, 2012.

\item Conference and workshop reviewing:
  \href{http://conf.researchr.org/home/pldi2015/}{PLDI 2015};
  \href{http://icfpconference.org/icfp2013/}{ICFP 2013};
  \href{http://ppopp2013.ics.uci.edu/}{PPoPP 2013};
  \href{http://research.microsoft.com/en-us/um/people/nswamy/plpv12/}{PLPV
    2012}.
\end{itemize*}

\subsection*{Departmental service}

\begin{itemize*}
\item Graduate Education Committee, Indiana University Computer
  Science Program, 2013--2014.
  
\item Website and mailing list administrator,
  \href{http://lambda.soic.indiana.edu}{Indiana University Programming
    Languages Group}, 2010--2014.

\item Officer, \href{http://csclub.soic.indiana.edu}{Indiana
  University Computer Science Club}, 2011--2013.

\item Organizer, \href{http://lambda.soic.indiana.edu/talks}{Indiana
  University PL Colloquium Series}, 2010--2012.  Coordinated speakers
  for weekly talk series.

\item Co-organizer and program committee member,
  \href{http://www.cs.indiana.edu/inwic/}{Indiana Celebration of Women
    in Computing (InWIC) 2012}.

\item President, Indiana University Computer Science Graduate
  Student Association, 2010--2011.

\item Steering Committee member,
  \href{http://wic.soic.indiana.edu/}{Indiana University Women in
    Informatics and Computing}, 2010--2011.
\end{itemize*}

\subsection*{Outreach activities}

\begin{itemize*}
\item Co-founder and co-organizer,
  \href{http://www.bangbangcon.com}{!!Con}, 2014--present.  !!Con is
  an annual volunteer-run conference consisting of ten-minute talks on
  the joy, excitement, and surprise of programming.

\item Program committee member,
  \href{http://tinytocs.org/vol3/}{\emph{Tiny Transactions on Computer
      Science} volume 3}, 2015. Tiny ToCS is the premier venue for
  peer-reviewed computer science research of $\leq$140 characters.

\item \href{http://www.recurse.com/residents}{Resident} at the
  \href{http://www.recurse.com}{Recurse Center}, New York, NY, summer
  2013, fall 2013, winter 2014.  The Recurse Center is a free,
  self-directed, educational retreat for programmers.
\end{itemize*}

\section*{Open source software contributions}
\begin{itemize*}

\item Contributor to
  \href{http://github.com/IntelLabs/RiverTrail}{River Trail}, a
  library, JIT compiler, and web browser extension to enable parallel
  programming in JavaScript, September 2014--February 2015.

\item Contributor to
  \href{http://hackage.haskell.org/package/lvish}{LVish}, the Haskell
  library for deterministic and quasi-deterministic parallel
  programming based on my dissertation work on LVars, February
  2013--present.

\item Contributor to the first ten releases of
  \href{http://rust-lang.org}{the Rust programming language},
  %% (\href{http://mail.mozilla.org/pipermail/rust-dev/2012-January/001256.html}{0.1},
  %% \href{http://mail.mozilla.org/pipermail/rust-dev/2012-March/001511.html}{0.2},
  %% \href{http://mail.mozilla.org/pipermail/rust-dev/2012-July/002087.html}{0.3},
  %% \href{http://mail.mozilla.org/pipermail/rust-dev/2012-October/002489.html}{0.4},
  %% \href{http://mail.mozilla.org/pipermail/rust-dev/2012-December/002787.html}{0.5},
  %% \href{http://mail.mozilla.org/pipermail/rust-dev/2013-April/003427.html}{0.6},
  %% \href{http://mail.mozilla.org/pipermail/rust-dev/2013-July/004667.html}{0.7},
  %% \href{http://mail.mozilla.org/pipermail/rust-dev/2013-September/005804.html}{0.8},
  %% \href{https://mail.mozilla.org/pipermail/rust-dev/2014-January/007753.html}{0.9},
  %% \href{https://mail.mozilla.org/pipermail/rust-dev/2014-April/009387.html}{0.10})
  as well as various pre-release versions, 2011--2014.  Contributions
  include work on extending the Rust typeclass system to support
  Haskell-style default methods; integer-literal suffix inference; and
  the self-dispatch, object extension, and method overriding features
  of the object system in a pre-release version of the language.
\end{itemize*}

\section*{Awards and fellowships}

\begin{itemize*}

\item
  Indiana University
  \href{http://www.indiana.edu/~grdschl/graduate-women-in-science.php}{Graduate
    Women in Science Fellowship}, 2008--2009.

\item National Merit Scholarship, 2000--2004.

\item Travel funding:
  \begin{itemize*}
  \item
    \href{http://www.doc.ic.ac.uk/~gds/PLMW/}{PLMW 2013} and
    \href{http://www.cis.upenn.edu/~sweirich/plmw12/}{PLMW 2012}
    travel awards for POPL, January 2013, January 2012.

  \item \href{http://cra-w.org/gradcohort}{CRA-W Graduate Cohort
    Workshop}, invitations and travel awards, March 2010, March 2009.

  \item
    \href{http://googleforstudents.blogspot.com/2008/12/google-workshop-for-women-engineers.html}{Google
      Workshop for Women Engineers}, invitation and travel award,
    January 2009.
  \end{itemize*}

\end{itemize*}

\section*{Other activities}

\begin{itemize*}

\item Competed (with Recurse Center staff and students) as {\em Hacker
  School batch[6]} in the \href{http://www.icfpcontest.org}{ICFP
  Programming Contest}, 2013.

\item Competed (with Alex Rudnick) as {\em Team K\&R} in the
  \href{http://www.icfpcontest.org}{ICFP Programming Contest}, 2011,
  2009, 2008.

\item Member of the
  \href{http://www.music.indiana.edu/departments/ensembles/cve.shtml}{Contemporary
    Vocal Ensemble}, Indiana University Jacobs School of Music, 2009.

\item Member of the
  \href{http://www.music.indiana.edu/departments/ensembles/university-chorale.shtml}{University
    Chorale}, Indiana University Jacobs School of Music, 2008--2009.

\item Member of the
  \href{http://www.grinnell.edu/academic/music/ensembles/singers}{Grinnell
    Singers}, Grinnell College Department of Music, 2000--2004.

\item Completed seven marathons since 2004, most recently in May 2012.

\end{itemize*}

\end{document}
